---
title: "Compare CommonMind Consortium signatures"
subtitle: 'hevir on biological data'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{Compare CommonMind Consortium signatures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---


<!---

cd /Users/gabrielhoffman/workspace/repos/heivr_analysis

cd ~/work/heivr_analysis

# rm -rf cmc_cache

ml git
R

system("git pull"); rmarkdown::render("cmc.Rmd");

https://hoffmg01.u.hpc.mssm.edu/work/heivr_analysis/cmc.html

scp sklar1:~/www/work/heivr_analysis/cmc.html .

--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  cache = TRUE,
  package.startup.message = FALSE,
  cache.lazy = FALSE)
```

```{r load, cache=FALSE}
library(heivr)
library(variancePartition)
library(parallel)
library(broom)
library(tidyverse)
```

```{r analysis, fig.width=4.5}
library(synapser)
synLogin()

# Load data and results
METADATA = readRDS(synGet('syn24180850')$path)
vobj.lst = readRDS(synGet('syn24180852')$path)
fitList = readRDS(synGet('syn24180854')$path)

# run top table for each cohort
tabList = lapply( fitList, function(fit){

	topTable(fit, coef="DxSCZ", number=Inf)
})

# merge results
df = merge(tabList[[1]], tabList[[2]], by="row.names")

# Pearson
with(df, cor(logFC.x, logFC.y))

# MLE
hobj = with(df, heivr(logFC.x, logFC.y, (logFC.x/t.x)^2, (logFC.y/t.y)^2))

hobj 
 
plot(hobj, xlab="logFC (MSSM-Penn-Pitt)", ylab="logFC (NIMH-HBCC)")
 
# MOM
hobj.mom = with(df, heivr(logFC.x, logFC.y, (logFC.x/t.x)^2, (logFC.y/t.y)^2, method="mom"))

hobj.mom
```


 
plot(hobj, xlab="logFC (MSSM-Penn-Pitt)", ylab="logFC (NIMH-HBCC)", showLines=TRUE, showShrinkage=TRUE, 
      showError=FALSE, length.arrow=.06, main="Original & posterior values")

# Ramp up analysis
Subsample subjects and get empirical standard errors.  Need to do this for simulation instead of theoretical se.

```{r de.rampup, cache=TRUE}
get_rho_rampup = function( coef, fractions=seq(.2, 1, length.out=3), nreps=10, BPPARAM = SnowParam(4, progressbar=TRUE) ){

  form.lst = list(
    `MSSM-Penn-Pitt`  = ~ Dx + Reported_Gender + RIN +scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + Institution*(ageOfDeath + cellFrac_ilr_1 + cellFrac_ilr_2 + cellFrac_ilr_3),
    `NIMH-HBCC`       = ~ Dx + Reported_Gender + RIN + scale(IntronicRate) + scale(IntragenicRate) + scale(IntergenicRate) + scale(rRNARate) + ageOfDeath + cellFrac_ilr_1 + cellFrac_ilr_2 + cellFrac_ilr_3)

  res = bplapply(1:nreps, function(k, vobj.lst, METADATA){
    suppressPackageStartupMessages({
    library(broom)
    library(tidyverse)
    library(heivr)
    library(RhpcBLASctl)
    library(dreamlet)
    library(variancePartition)})
    omp_set_num_threads(1)

    res = lapply( fractions, function(frac){

      fit_marginal.lst = lapply( vobj.lst, function(vobj){
        i = match(colnames(vobj), rownames(METADATA))
        info = METADATA[i,]

        # bootstrap using replacement
        idx = sample.int( ncol(vobj), ncol(vobj)*frac, replace=TRUE)

        info.sub = droplevels(info[idx,])

        form = update(form.lst[[info$Cohort[1]]], ~ . - Dx)
        dsgn = model.matrix(form, info.sub)
        dcmp = svd(dsgn)
        incl = dcmp$d > 1e-6
        data2 = with(dcmp, u[,incl,drop=FALSE] %*% diag(d[incl]))
        data2 = data.frame(data2)
        form_mod = paste0('~ Dx + ', paste0('X', 1:ncol(data2), collapse=" + ")) 

        tryCatch({
          fit = dream(vobj[,idx], form_mod, info.sub, quiet=TRUE)  
          fit = eBayes( fit )
        }, error = function(e) fit = NULL, 
        warning = function(w) fit = NULL)
        })
      names(fit_marginal.lst) = names(vobj.lst)

      if( any(sapply(fit_marginal.lst, is.null)) ) return( NULL)

      # run top table for each cohort
      tabList = lapply( fit_marginal.lst, function(fit){
        topTable(fit, coef=coef, number=Inf)
      })

      # merge results
      df = merge(tabList[[1]], tabList[[2]], by="row.names")

      # Pearson
      fit.Pearson = with(df, cor.test(logFC.x, logFC.y)) %>% tidy %>% mutate(se = sqrt((1 - estimate^2)/parameter))

      # MLE
      hobj = with(df, heivr(logFC.x, logFC.y, (logFC.x/t.x)^2, (logFC.y/t.y)^2, LRT=FALSE))

      # MOM
      hobj.mom = with(df, heivr(logFC.x, logFC.y, (logFC.x/t.x)^2, (logFC.y/t.y)^2, method="mom", nboot=0))

      data.frame( fraction = frac, 
                  k = k,
                  rho.Pearson = fit.Pearson$estimate,
                  rho.heivr = hobj$rho,
                  rho.mom = hobj.mom$rho)
    })
    do.call(rbind, res)
  }, vobj.lst=vobj.lst, METADATA=METADATA, BPPARAM=BPPARAM)
  res = do.call(rbind, res)

  # combine results
  res = rbind( with(res, data.frame( 
                      Method = "heivr",
                      fraction = fraction, 
                      rho = rho.heivr)),
                with(res, data.frame( 
                      Method = "MOM",
                      fraction = fraction, 
                      rho = rho.mom)),
                with(res, data.frame( 
                      Method = "Pearson",
                      fraction = fraction, 
                      rho = rho.Pearson)) )

  res$Method = factor(res$Method, c("heivr", "MOM", "Pearson"))
  res$coef = coef

  res
}
```


```{r run_ramp}
# ramp up analysis
coefTest = c('DxSCZ', 'Reported_GenderMale', 'ageOfDeath')

df.ramp = lapply(coefTest, function(coef){
  get_rho_rampup(coef, c(0.05, seq(.1, 1, length.out=3)), nreps=20, BPPARAM = SnowParam(16))
  })
df.ramp = do.call(rbind, df.ramp)
```

Note slight x-axis offset was added to plot so confidence intervals would be visible
```{r rampup.plot.SCZ, cache=FALSE, fig.width=10, fig.height=5}
res2 = df.ramp %>%
  group_by(coef, Method, fraction) %>%
  summarize( rho.mean = mean(rho), rho.se = sd(rho))

res2$offset = sapply(as.character(res2$Method), function(x) 
                switch(x, Pearson=-0.01, heivr = 0, MOM=0.01))

ymin = min(with(res2, pmax(-1,rho.mean - 1.96*rho.se)))

ggplot(res2, aes(fraction + offset, rho.mean, color=Method)) +
    geom_hline(yintercept=0, color="black", linetype="dashed") +
    geom_point() + 
    geom_errorbar(aes(ymin = pmax(-1,rho.mean - 1.96*rho.se),
                       ymax = pmin(1, rho.mean + 1.96*rho.se)),
                  width=0) +
    theme_bw() +
    theme(aspect.ratio=1) +
    scale_x_continuous(limits=c(0,1.02), expand=c(0,0), 
      labels=c('0', '0.25', '0.50', '0.75', '1')) +
    scale_y_continuous(limits=c(ymin - 0.04,1.04), expand=c(0,0)) +
    xlab("Fraction of subjects included from each study") +
    ylab("Estimated correlation") +
    scale_color_manual(values=c("red3", "green3", "blue3")) +
    facet_wrap(~coef)
```







