---
title: "Compare AD signatures"
subtitle: 'PsychAD vs Mathys, et al'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
---


<!---

cd /Users/gabrielhoffman/workspace/repos/heivr_analysis

cd ~/work/heivr_analysis

# rm -rf cmc_cache

ml git
R

system("git pull"); rmarkdown::render("PsychAD_Mathys.Rmd");

https://hoffmg01.u.hpc.mssm.edu/work/heivr_analysis/PsychAD_Mathys.html


--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  cache = TRUE,
  package.startup.message = FALSE,
  cache.lazy = FALSE)
```

```{r load, cache=FALSE}
library(dreamlet)
library(heivr)
library(tidyverse)

# file = "/sc/arion/projects/CommonMind/hoffman/dreamlet_analysis/PsychAD_r0/PsychAD_dreamlet.RDS"
file = "~/Downlaods/PsychAD_dreamlet.RDS"
fit.psychad = readRDS(file)
tab.psychad = topTable(fit.psychad, coef="DxAD", number=Inf) %>%
                as_tibble %>%
                mutate(se = logFC / t)

# file = "/sc/arion/projects/CommonMind/hoffman/eval_methods/dreamlet/Mathys_dreamlet.RDS"
file = "~/Downloads/Mathys_dreamlet.RDS"
fit.mathys = readRDS(file)
tab.mathys = topTable(fit.mathys, coef="educ", number=Inf) %>%
                as_tibble %>%
                mutate(se = logFC / t)



pair = c(psychad = "Astro", mathys = "Ast")

tab.psychad.sub = tab.psychad %>% 
        filter(assay == pair[['psychad']])

tab.mathys.sub = tab.mathys %>% 
        filter(assay == pair[['mathys']])

tab = inner_join(tab.psychad.sub, tab.mathys.sub, by="ID")


with(tab, cor(logFC.x, logFC.y))

hobj.mom = with(tab, heivr(logFC.x, logFC.y, se.x^2, se.y^2, LRT=FALSE, method="mom"))
hobj.mom


hobj.mle = with(tab, heivr(logFC.x, logFC.y, se.x^2, se.y^2, LRT=FALSE))
hobj.mle





plot(hobj.mle)

with(tab, errorFraction(logFC.x, se.x^2))
with(tab, errorFraction(logFC.x, se.x^2, method="mom"))


with(tab, errorFraction(logFC.y, se.y^2))
with(tab, errorFraction(logFC.y, se.y^2, method="mom"))

n = 1000
se.x = rgamma(n, 6, 1)
se.y = rgamma(n, 6, 1)
x = rnorm(n, sd=se.x)
y = rnorm(n, sd=se.y)

heivr(x,y, se.x^2, se.y^2, method="mom")
heivr(x,y, se.x^2, se.y^2, LRT=FALSE)



```

# all pairs
```{r all}

pair = c(psychad = "Micro_PVM", mathys = "Ast")

tab.psychad.sub = tab.psychad %>% 
        filter(assay == pair[['psychad']])

df = lapply( unique(tab.mathys$assay), function(cellType){

  tab.mathys.sub = tab.mathys %>% 
          filter(assay == cellType)

  tab = inner_join(tab.psychad.sub, tab.mathys.sub, by="ID")

  df = list()
  df$cor = with(tab, cor(logFC.x, logFC.y))

  hobj.mle = with(tab, heivr(logFC.x, logFC.y, se.x^2, se.y^2, LRT=FALSE))
  df$cor.heivr = hobj.mle$rho
  df$cor.heivr.se = hobj.mle$rho.se

  hobj.mom = with(tab, heivr(logFC.x, logFC.y, se.x^2, se.y^2, LRT=FALSE, method="mom"))
  df$cor.mom = hobj.mom$rho
  df$cor.mom.se = hobj.mom$rho.se

  data.frame(cellType, df)
})
df = do.call(rbind, df)





